# Continuous Random Variables {#continuous}

## Motivating Example {-}

In a Poisson process, _the number of arrivals_ on any interval is 
a random variable that follows a Poisson distribution. 
But what about _the time of the first arrival_? This is also 
a random variable, but it is a different kind of random variable. 
The time of the first arrival does not have to be an integer. It 
can happen at $1.2$ seconds, $2.173$ seconds, $\sqrt{2}$ seconds, or 
$\pi$ seconds. In fact, any real number from $0$ to $\infty$ is a 
possible outcome of this random variable. As we will see in this lesson, 
this has massive consequences for the random variable.

Suppose radioactive particles hit a Geiger counter at a rate of 
$\lambda = 0.8$ particles per second. We use $T$ to denote the time 
of the first arrival---that is, the time that the first particle 
hits the Geiger counter. We can calculate probabilities involving $T$. 
For example, we can calculate the probability that the first particle 
arrives after 2 seconds by rewriting this event in terms of the 
number of arrivals on an interval:
\[ P(T > 2) = P(\text{0 particles between 0 and 2 seconds})  \]
We know that the number of particles between 0 and 2 seconds 
follows a $\text{Poisson}(\mu=0.8 \cdot 2)$ distribution, so we 
just need to evaluate the p.m.f. at $x=0$:
\begin{equation} 
P(T > 2) = e^{-0.8 \cdot 2} \frac{(0.8 \cdot 2)^0}{0!} = e^{-1.6} \approx .202. 
(\#eq:t-gt-2)
\end{equation}

We can also calculate the probability that the first arrival happens between 
$2$ and $3$ seconds, as:
\[ P(2 < T < 3) = P(T > 2) - P(T > 3). \]
We calculate $P(T > 3)$ in much the same way as we calculated $P(T > 2)$ above:
\begin{equation}
P(T > 3) = e^{-0.8 \cdot 3} \frac{(0.8 \cdot 3)^0}{0!} = e^{-2.4} \approx .091. 
(\#eq:t-gt-3)
\end{equation}
So we see that 
\[ P(2 < T < 3) = e^{-1.6} - e^{-2.4} \approx .111. \]

We calculated two specific probabilities, but what about the distribution of $T$?
Here are some questions we will answer in this lesson:
- What is the c.d.f. of $T$?
- Does $T$ have a p.m.f.?

## Theory {-}

First, let's calculate the c.d.f. of $T$. 
```{example geiger-cdf, name="The CDF of the First Arrival"}
Remember that the c.d.f. is defined as $F(t) = P(T \leq t)$ as a function of $t$. 
First, we use the complement rule:
\[ F(t) = P(T \leq t) = 1 - P(T > t). \]
Now, we calculate $P(T > t)$, in much the same way that we calculated $P(T > 2)$ 
  in \@ref(eq:t-gt-2) and $P(T > 3)$ in \@ref(eq:t-gt-3):
\[ P(T > t) = e^{- 0.8 \cdot t} \frac{(0.8 \cdot t)^0}{0!} = e^{-0.8 t}. \]
This formula works for $t \geq 0$. Since times cannot be negative, we have 
the following formula for $t < 0$:
\[ P(T > t) = 1 \]

Putting everything together, the c.d.f. of $T$ is 
\[ F(t) = 1 - P(T > t) = \begin{cases} 1 - e^{-0.8 t} & t \geq 0 \\ 0 & t < 0 \end{cases}. \]
This function is graphed below.
```{r t-cdf-graph, echo=FALSE, fig.show = "hold", fig.align = "default",fig.cap="CDF of the First Arrival"}
t <- seq(0, 5.5, by=0.01)
plot(t, 1 - exp(-0.8 * t), type="l", lwd=3, col="orange",
     xaxt="n", yaxt="n", bty="n",
     xlab="t", ylab="Cumulative Distribution Function F(t)", 
     xlim=c(-0.5, 5.5), ylim=c(-0.05, 1.05))
lines(c(-0.5, 0), c(0, 0), lwd=3, col="orange")
axis(1, pos=0)
axis(2, pos=0)
```

Notice how different this c.d.f. is, compared with the ones we graphed in 
Lesson \@ref(cdf). In Figure \@ref(fig:cdf-graph), the c.d.f. was a 
step function, with a jump at each possible value. By contrast, the 
c.d.f. above is continuous. This is the main distinction between the 
kinds of random variables we were studying before, which are called 
**discrete random variables**, and the kinds of 
random variables we study in this lesson, which are called 
**continuous random variables**.

```{definition, name="Continuous Random Variable"}
A random variable is called **continuous** if its c.d.f. is a 
continuous function.
```

With discrete random variables, we can visualize the shape of the 
distribution by graphing its p.m.f. Recall that the p.m.f. specified the 
probability that the random variable is _equal to_ $x$. Continuous random 
variables do not have a p.m.f. because the probability of any exact outcome is zero.

```{example, name="Probability of an Exact Outcome"}
Continuing with Example \@ref(exm:geiger-cdf), what is the probability that the 
first particle hits the Geiger counter at _exactly_ 2 seconds 
(that is, 2.0000000... seconds)?

First, consider the probability that the first arrival is in the 
interval $(2 - \epsilon, 2 + \epsilon)$, where $\epsilon$ is a small 
positive number. No matter how small an $\epsilon$ we choose, this 
probability will always be greater than the probability that the 
first arrival happens at _exactly_ 2 seconds, since this interval includes 2 seconds, 
as well as other outcomes. (For example, if $\epsilon = 0.05$, then this interval would 
also include the possibility that the first arrival happens at 1.98 seconds or 
2.01 seconds.) 

The number of arrivals on the interval $(2 - \epsilon, 2 + \epsilon)$ is a 
Poisson random variable with parameter $\mu = 0.8 \cdot 2\epsilon$. Therefore, 
the probability that there are no arrivals on the interval is:
\begin{align*}
P(\text{at least 1 arrival on interval}) &= 1 - P(\text{no arrivals on interval}) \\
&= 1 - e^{-0.8 \cdot 2\epsilon} \frac{(0.8 \cdot 2 \epsilon)^0}{0!} \\
&= 1 - e^{-1.6 \epsilon}.
\end{align*}
But $\epsilon$ can be arbitrarily small. As $\epsilon \to 0$, this probability approaches 0. 

Therefore, the probability that the first arrival happens at _exactly_ 2 seconds is zero.
```

Because the probability of every outcome in a continuous random variable is zero, 
continuous random variables cannot be described by their p.m.f. Instead, they are 
described by a similar function called the **probability density function**.

```{definition pdf, name="Probability Density Function"}
The **probability density function** (or p.d.f.) of a continuous random variable 
is defined to be the derivative of the c.d.f.
\[ f(x) \overset{\text{def}}{=} F'(x). \]
The values of $f(x)$ do not represent the probability that the random variable 
is equal to $x$ (because for a continuous random variable, that probability is 
always equal to 0).
```

```{example geiger-pdf, name="The PDF of the First Arrival"}
Continuing with Example \@ref(exm:geiger-cdf), the c.d.f. was calculated to be 
\[ F(t) = \begin{cases} 1 - e^{-0.8 t} & t \geq 0 \\ 0 & t < 0 \end{cases}. \]
Taking the derivative, we have 
\begin{equation}
f(t) = F'(t) = \begin{cases} 0.8 e^{-0.8 t} & t \geq 0 \\ 0 & t < 0 \end{cases}. 
(\#eq:geiger-pdf)
\end{equation}
This p.d.f. is graphed below.

```{r t-pdf-graph, echo=FALSE, fig.show = "hold", fig.align = "default",fig.cap="PDF of the First Arrival"}
t <- seq(0, 5.5, by=0.01)
plot(t, 0.8 * exp(-0.8 * t), type="l", lwd=3, col="blue", 
     xaxt="n", yaxt="n", bty="n",
     xlab="t", ylab="Probability Density Function f(t)", 
     xlim=c(-0.5, 5.5), ylim=c(-0.05, 0.81))
axis(1, pos=0)
axis(2, pos=0)
lines(c(-0.5, 0), c(0, 0), lwd=3, col="blue")
```

This p.d.f. tells us that the first arrival is more likely to happen sooner, 
rather than later. However, the values of the p.d.f. are not easy to interpret. 
They do not represent probabilities, but rather probability _densities_. 

If the values of the p.d.f. do not represent probabilities, how do we calculate 
probabilities using the p.d.f.? It turns out that _areas under the p.d.f._ 
represent probabilities.

```{theorem calculating-probs-pdf, name="Calculating Probabilities Using the PDF"}
Let $X$ be a continuous random variable with p.d.f. $f(x)$. Then:
\[ P(a < X \leq b) = \int_a^b f(x)\,dx. \]
```{proof}
Let $F(x)$ be the c.d.f. of $X$. Then, we know that 
\[ P(a < X \leq b) = P(X \leq b) - P(X \leq a) = F(b) - F(a). \]
By the [Fundamental Theorem of Calculus](https://en.wikipedia.org/wiki/Fundamental_theorem_of_calculus#Second_part), 
\[ F(b) - F(a) = \int_a^b f(x)\,dx. \]
```

Let's recalculate the probabilities using the p.d.f. \@ref(eq:geiger-pdf).

```{example, name="Calculating Probabilities Using the PDF"}
We know that the p.d.f. of the first arrival, $T$, is 
\[ f(t) = \begin{cases} 0.8 e^{-0.8 t} & t \geq 0 \\ 0 & t < 0 \end{cases}. \]
To calculate probabilities using Theorem \@ref(thm:calculating-probs-pdf), we 
integrate the p.d.f. 

For example, the probability that the first arrival happens between 2 and 3 seconds 
is:
\[ P(2 < T < 3) = \int_2^3 f(t)\,dt = \int_2^3 0.8 e^{-0.8 t}\,dt = .111, \]
and the probability that the first arrival happens after 2 seconds is 
\[ P(T > 2) = \int_2^\infty f(t)\,dt = \int_2^\infty 0.8 e^{-0.8 t}\,dt = .202.\]
The areas that these probabilities represent are shown on the graphs below.
```{r, echo=FALSE, fig.show = "hold", fig.align = "default",fig.cap="PDF of the First Arrival", fig.asp=1.2}
par(mfrow=c(2, 1))
t <- seq(0, 5.5, by=0.01)
plot(t, 0.8 * exp(-0.8 * t), type="l", lwd=3, col="blue",
     main="P(2 < T < 3)",
     xaxt="n", yaxt="n", bty="n",
     xlab="t", ylab="Probability Density Function f(t)", 
     xlim=c(-0.5, 5.5), ylim=c(-0.05, 0.81))
axis(1, pos=0)
axis(2, pos=0)
range <- seq(2, 3, by=0.01)
polygon(c(2, range, 3), c(0, 0.8 * exp(-0.8 * range), 0), 
        border=NA, col=rgb(1, 0, 0, 0.5))
lines(c(-0.5, 0), c(0, 0), lwd=3, col="blue")

plot(t, 0.8 * exp(-0.8 * t), type="l", lwd=3, col="blue",
     main="P(T > 2)",
     xaxt="n", yaxt="n", bty="n",
     xlab="t", ylab="Probability Density Function f(t)", xlim=c(-0.5, 5.5), ylim=c(-0.05, 0.81))
axis(1, pos=0)
axis(2, pos=0)
range <- seq(2, 5.5, by=0.01)
polygon(c(2, range, 5.5), c(0, 0.8 * exp(-0.8 * range), 0), 
        border=NA, col=rgb(1, 0, 0, 0.5))
lines(c(-0.5, 0), c(0, 0), lwd=3, col="blue")
```

What about the probability that the first arrival happens before 3 seconds? 
Technically, we integrate the p.d.f. from $-\infty$ to $3$. However, $f(t) = 0$ 
for $-\infty < t < 0$, so we effectively integrate the p.d.f. from $0$ to $3$:
\[ P(T < 3) = \int_{-\infty}^3 f(t)\,dt = \int_0^3 0.8 e^{-0.8 t}\,dt = .909. \]
```

So far, we saw that we take the derivative of the c.d.f. to get the p.d.f. Likewise, if we 
have the p.d.f., we can take its integral to get the p.d.f.
\begin{equation}
F(x) = \int_{-\infty}^x f(t)\,dt.
(\#eq:pdf-to-cdf)
\end{equation}

```{example, name="(Re)deriving the CDF of the First Arrival"}
Suppose we only had the p.d.f. of the first arrival $T$:
\[ f(t) = \begin{cases} 0.8 e^{-0.8 t} & t \geq 0 \\ 0 & t < 0 \end{cases}. \]

To get the c.d.f., we would integrate. Since the p.d.f. is 0 when $t < 0$, the integral 
effectively starts from $t=0$.
\[ F(x) = \int_{-\infty}^x f(t)\,dt = \int_0^x 0.8 e^{-0.8 t}\,dt = 1 - e^{-0.8 x}. \]
(This integral can be calculated using the $u$-substitution $u=0.8 t$, or [using 
Wolfram Alpha](https://www.wolframalpha.com/input/?i=integrate+0.8*exp%28-0.8*t%29+from+t%3D0+to+x).)
```


## Essential Practice {-}

1. Packets arrive at a certain node on the university’s intranet at 10 packets per minute, on average.
Assume packet arrivals meet the assumptions of a Poisson process.

    a. What is the p.d.f. of $T$, the time of the first arrival?
    b. What is the probability that the first arrival happens between 10 seconds and 30 seconds? 
    (Note that the rate in the problem is given in _minutes_.)

2. Two Cal Poly students, Ferris and Cameron, are frequently late to class. Cal Poly 
classes start at 10 minutes past the hour mark.

    Let $X$ be the time (in minutes) that Ferris arrives at class after the hour mark. The p.d.f. 
    of $X$ is 
    \[ f_X(x) = \begin{cases} \frac{1}{60} & 0 \leq x < 60 \\ 0 & \text{otherwise} \end{cases}.  \]
    Let $Y$ be the time (in minutes) that Cameron arrives at class after the hour mark. The p.d.f. 
    of $Y$ is 
    \[ f_Y(y) = \begin{cases} \frac{60 - y}{1800} & 0 \leq y < 60 \\ 0 & \text{otherwise} \end{cases}.  \]

    a. Sketch a graph of the two p.d.f.s. Without doing any calculations, who is more likely to arrive 
    on time (i.e., within the first 10 minutes of the hour)?
    b. Determine the c.d.f.s of $X$ and $Y$. 
    c. Calculate the probability that Ferris arrives on time. You should be able to calculate 
    this probability in three ways: (1) finding the area under the p.d.f. using geometry, (2) 
    finding the area under the p.d.f. using calculus, and (3) using the c.d.f.
    d. Calculate the probability that Cameron arrives on time. You should also be able to calculate 
    this probability in three ways.

3. The distance (in hundreds of miles) driven by a trucker in one day is a continuous
random variable $X$ whose cumulative distribution function (c.d.f.) is given by: 
\[ F(x) = \begin{cases} 0 & x < 0 \\ x^3 / 216 & 0 \leq x \leq 6 \\ 1 & x > 6 \end{cases}. \]

    a. Determine the p.d.f. Sketch a graph.
    b. Calculate the probability that the trucker travels more than 5 miles in a day. (You should 
    be able to calculate this in at least 2 ways: directly from the c.d.f. and using the p.d.f. 
    that you calculated in the previous part.)
    c. Calculate the probability that the trucker travels exactly 200 miles in a day.
    
4. Let $T$ be how late that a professor lets her class out (in minutes). 
(If $T$ is negative, then she finishes the class early.) The p.d.f. of $T$ is 
\[ f(t) = \begin{cases} c(2 + t) & -2 < t < 0 \\ c(2 - t) & 0 \leq t < 2 \\ 0 & \text{otherwise} \end{cases}, \]
where $c$ is a constant.

    a. Determine the value of $c$ necessary to make this a proper p.d.f. (_Hint:_ What does the 
    total probability have to be?) Sketch the p.d.f.
    b. Calculate the probability that the professor finishes between 1 minute early and 
    30 seconds late, i.e., $P(-1 < T < 0.5)$.
