# Double Counting {#double-counting}

## Motivating Example {-}

The French nobleman (and avid gambler) Chevalier de Méré knew that betting on at least one 
six (⚅) in 4 rolls of a die was a favorable bet for him. Once other gamblers caught on, he 
devised a new bet: at least one double-six (⚅⚅) in 24 rolls of two dice. Although he did not know 
how to calculate the probabilities, he reasoned that the two bets should be equivalent, since 

- double-sixes are $1/6$ as likely as a single six,
- but there are $6$ times as many rolls to compensate

Are the two bets equivalent?

## Theory {-}

Here is a common (but wrong) way of calculating the probability of getting at least one six in 
4 rolls of a die.

\begin{align*}
P(\text{at least one ⚅}) &= P(\text{⚅ on 1st roll, or ⚅ on 2nd roll, or ⚅ on 3rd roll or ⚅ on 4th roll}) \\
&= P(\text{⚅ on 1st roll}) + P(\text{⚅ on 2nd roll}) + P(\text{⚅ on 3nd roll}) + P(\text{⚅ on 4th roll}) \\
&= \frac{1}{6} + \frac{1}{6} + \frac{1}{6} + \frac{1}{6} \\
&= \frac{4}{6} ?
\end{align*}

What's wrong with the above reasoning? The problem is that 
\[ P(A \text{ or } B) \neq P(A) + P(B) \]
in general. The reason is that $P(A) + P(B)$ double counts outcomes where $A$ and $B$ both happen.

For example, suppose $A$ is "⚅ on 1st roll" and $B$ is "⚅ on 2nd roll". Then, it is not hard to see 
that the event $A \text{ or } B$ happens on 11 out of 36 outcomes, so 
\[ P(A \text{ or } B) = \frac{11}{36} \neq \frac{12}{36} = \frac{1}{6} + \frac{1}{6} = P(A) + P(B). \]
$P(A) + P(B)$ double counts the outcome where _both_ rolls were ⚅s.

```{r, echo=FALSE, engine='tikz', out.width='60%', fig.ext='png', fig.align='center', fig.cap='Double Counting Dice Rolls', engine.opts = list(template = "tikz_template.tex")}
\begin{tikzpicture}
	
	\foreach \i 	in {1, 2, 3, 4, 5, 6} {
		\fill [blue!50,opacity=.3] (1.5*6 - .75, -1*\i - .5) rectangle (1.5*6 + .75, -1*\i + .5);
	}
	\foreach \i 	in {1, 2, 3, 4, 5, 6} {	
		\fill [red!50,opacity=.3] (1.5*\i - .75, -1*6 - .5) rectangle (1.5*\i + .75, -1*6 + .5);
	}

	\foreach \i in {1, ..., 6} {
		\foreach \j in {1, ..., 6} {
			\node[rectangle,draw] at (1.5*\i, -\j) {\Large\epsdice{\j}\ \epsdice{\i}};
		}
	}

    		\foreach \i 	in {1, ..., 6} {
    			\foreach \j in {1, ..., 6} {
    				\node[rectangle,draw,opacity=.3] at (1.5*\i, -\j) {\Large\epsdice{\j}\ \epsdice{\i}};
    			}
    		}
		
		\foreach \i 	in {1, ..., 6} {
			\node[rectangle,draw] at (1.5*\i, -1*4) {\Large\epsdice{4}\ \epsdice{\i}};
			\node[rectangle,draw] at (1.5*4, -1*\i) {\Large\epsdice{\i}\ \epsdice{4}};
		}
\end{tikzpicture}
```

One way to avoid double counting is to subtract the cases that are double counted.

```{theorem, name="Inclusion-Exclusion Principle"}
\[ P(A \text{ or } B) = P(A) + P(B) - P(A \text{ and } B) \]
```

```{r, echo=FALSE, engine='tikz', out.width='60%', fig.ext='png', fig.align='center', fig.cap='Intuition for the Inclusion-Exclusion Principle'}

\tikzset{filled/.style={fill=circle area, draw=circle edge, thick},
    outline/.style={draw=circle edge, thick}}

\begin{tikzpicture}
    \draw[fill=red!50,fill opacity=0.5] (0, 0) circle (30pt);
    \draw[fill=blue!50,fill opacity=0.5] (1, 0) circle (30pt);

    \node at (-0.4, 0) {{\scriptsize $A$}};
    \node at (1.4, 0) {{\scriptsize $B$}};
    \node at (0.5, 0) {{\scriptsize $A \text{ and } B$}};
\end{tikzpicture}
```

So, for example:
\begin{align*} 
P(\text{at least one ⚅ in 2 rolls}) &= P(\text{⚅ on 1st roll}) + P(\text{⚅ on 2nd roll}) - P(\text{⚅ on both rolls}) \\
&= \frac{1}{6} + \frac{1}{6} - \frac{1}{36} \\
&= \frac{11}{36}
\end{align*}

However, this approach does not scale well to calculating the probability of at least one ⚅ in 4 rolls. 

In many situations, it is easier to calculate the probability that an event does _not_ happen, also known as the 
**complement** of the event. Because the total probability has to be 1, the two probabilities are related by the following formula.

```{theorem, name="Complement Rule"}
\[ P(\text{not } A) = 1 - P(A). \]
```

Let's apply the Complement Rule to the Chevalier de Méré's problem. To calculate the probability of getting at least one ⚅ in 4 rolls, 
we can calculate the probability of the complement. If we did not get at least one ⚅, that must mean that we got no ⚅s. This means that 
every roll was one of the other 5 outcomes. This probability is much easier to calculate using the counting tricks we have learned.
\begin{align*}
P(\text{at least one ⚅}) &= 1 - P(\text{no ⚅s}) \\
&= 1 - \frac{5^4}{6^4} \\
&\approx 0.5177.
\end{align*}


## Examples {-}

1. In poker, a "two pair" hand has 2 cards of one rank, 
2 cards of another rank, and 1 card of a third rank. For example, the hand 
2, 2, Q, Q, J is a "two pair". Your friend calculates the probability of "two pair" as follows:

    - There are $\binom{52}{5}$ equally likely hands (where order does not matter).
    - We count the number of ways to choose the first pair. There are $13$ choices for the rank and $\binom{4}{2}$ choices for the two cards within the rank, so there are $13 \times \binom{4}{2}$ ways.
    - Next, we count the ways to choose the second pair. Since one rank has already been chosen, there are $12 \times \binom{4}{2}$ ways to do this.
    - Finally, we choose the remaining card. There are $11 \times \binom{4}{1} = 44$ ways to do this.
    
    Your friend calculates the probability as 
    \[ \frac{13 \times \binom{4}{2} \times 12 \times \binom{4}{2} \times 44}{\binom{52}{5}} \approx .095, \] 
    but then [finds online](http://www.math.hawaii.edu/~ramsey/Probability/PokerHands.html) that the actual probability of "two pair" is only $.0475$. This number is exactly half the probability that your friend got, so he suspects that he double-counted. But where?

2. Complete the calculation for the Chevalier de Méré. Calculate the probability of 
getting at least one ⚅⚅ in 24 rolls of two dice.

