# Fourier Transforms {#fourier}

## Continuous-Time Fourier Transforms {-}

The Fourier transform is an alternative representation of a signal. It describes the frequency content 
of the signal. It is defined as 
\begin{equation}
G(f) \overset{\text{def}}{=} \int_{-\infty}^\infty g(t)e^{-j2\pi f t}\,dt.
(\#eq:fourier)
\end{equation}
Notice that it is a function of frequency $f$, rather than time $t$. Notice also that it is complex-valued, 
since its definition involves the imaginary number $j \overset{\text{def}}{=} \sqrt{-1}$.

The following video explains the visual intuition behind the Fourier transform. Note that this video 
uses $i$ to denote the imaginary number $\sqrt{-1}$, whereas we use $j$ (which is common in electrical engineering, 
to avoid confusion with current).

<iframe width="560" height="315" src="https://www.youtube.com/embed/spUNpyF58BY" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

To calculate the Fourier transform of a signal, we will rarely use \@ref(eq:fourier). Instead, we will 
look up the Fourier transform of the signal in a table like Appendix \@ref(ctft) and use properties of the 
Fourier transform (as shown in Appendix \@ref(fourier-properties)).

```{example}
Calculate the Fourier transform of $g(t) = 80e^{-20t} u(t)$. 
```

```{solution}
This signal is most similar to $e^{-t} u(t)$ in Appendix \@ref(ctft), whose Fourier transform is 
$\frac{1}{1 + j2\pi f}$. The only differences are:
  
- Our signal has an extra factor of 80 in front. This factor comes along for the ride by linearity of the Fourier transform.
- Our signal is time-scaled by a factor of 20, so we have to apply the scaling property. 
Since we _multiplied_ by 20 in the time domain, we have to _divide_ by 20 in the 
frequency domain, on both the _inside_ and the _outside_.

Putting everything together, we see that the Fourier transform is:
\[ G(f) = 80 \frac{1}{20} \frac{1}{1 + j 2\pi \frac{f}{20}}. \]

This is a complex-valued function. That is, at each frequency $f$, $G(f)$ is a complex number, with a 
real and an imaginary component. To make it easier to visualize, we calculate its magnitude 
using Theorem \@ref(thm:calculating-magnitude). 
\begin{align*}
|G(f)| = \sqrt{|G(f)|^2} &= \sqrt{G(f) \cdot G^*(f)} \\
&= \sqrt{\frac{4}{1 + j2\pi \frac{f}{20}} \cdot \frac{4}{1 - j2\pi \frac{f}{20}}} \\
&= \sqrt{\frac{4}{1 + (2\pi \frac{f}{20})^2}}
\end{align*}

Now, the _magnitude_ of the Fourier transform is a function we can easily visualize. 
From the graph, we see that the signal has more low-frequency content than high-frequency content.
```

```{r, echo=FALSE, fig.show = "hold", fig.align = "default"}
fs <- seq(-50, 50, by=.1)
plot(fs, sqrt(4 / (1 + (2*pi*fs/20)^2)), col="orange", type="l", lwd=2,
     xaxt="n", yaxt="n", bty="n",
     xlab="Frequency (f)", ylab="Fourier Transform Magnitude |G(f)|")
axis(1, pos=0)
axis(2, pos=0, at=(0:4) / 2)
```

## Discrete-Time Fourier Transforms {-}

Discrete-time signals are obtained by sampling a continuous-time signal 
at regular time intervals. The sampling rate $f_s$ (in Hz) specifies the 
number of samples per second. The signal below is sampled at a rate of 
$f_s = 16$ Hz.

```{r, echo=FALSE, fig.show = "hold", fig.align = "default"}
ts <- seq(0, 1, .001)
plot(ts, sin(2 * pi * 4 * ts) + sin(2 * pi * 6 * ts), type="l",
     xlab="Time (seconds)", ylab="Amplitude")

ts <- seq(0, 1, length.out=17)
points(ts, sin(2 * pi * 4 * ts) + sin(2 * pi * 6 * ts), pch=19)
```

We write $x[n] = x(t_n)$ for the $n$th time sample, where $t_n = \frac{n}{f_s}$.

It's possible to choose a sampling rate $f_s$ that is too low. In the graph below, 
the underlying continuous-time signal (in black) is a sinusoid with a frequency of 14 Hz. 
However, because the discrete-time signal (in red) was sampled at a rate of 16 Hz, the 
sinusoid appears to have a frequency of 2 Hz.

```{r, echo=FALSE, fig.show = "hold", fig.align = "default"}
ts <- seq(0, 1, .001)
plot(ts, sin(2 * pi * 14 * ts), type="l", col="black",
     xlab="Time (seconds)", ylab="Amplitude")
lines(ts, -sin(2 * pi * 2 * ts), col="red")

ts <- seq(0, 1, length.out=17)
points(ts, sin(2 * pi * 14 * ts), pch=19, col="red")
```

We say that the higher frequency is **aliased** by the lower one.

At a sampling rate of $f_s$, any frequencies outside the range
\[ (-f_s / 2, f_s / 2) \]
will be aliased by a frequency inside this range. This "maximum frequency" 
of $f_s / 2$ is known as the **Nyquist limit**.
	
Therefore, the Fourier transform of a discrete-time signal is effectively only 
	defined for frequencies from $-f_s/2$ to $f_s/2$. Contrast this with the Fourier transform of a 
	continuous-time signal, which is defined on all frequencies from $-\infty$ to $\infty$.
	
Aliasing is not just a theoretical problem. For example, if helicopter blades are spinning 
too fast for the frame rate of a video, then they can look like they are not spinning at all!

<iframe width="560" height="315" src="https://www.youtube.com/embed/AYQAKwCxScc" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
	
When calculating the Fourier transform of a discrete-time signal, we typically work with 
_normalized_ frequencies. That is, the frequencies are in units of "cycles per _sample_" instead of 
"cycles per _second_". (Another way to think about this is that we assume all discrete-time signals are 
sampled at 1 Hz.) As a result, the Nyquist limit is $1/2$, so the **Discrete-Time Fourier Transform** 
is defined only for $|f| < 0.5$.

\begin{align}
G(f) &\overset{\text{def}}{=} \sum_{n=-\infty}^\infty g[n] e^{-j2\pi f t}\,dt & -0.5 < f < 0.5
(\#eq:discrete-fourier)
\end{align}

To calculate the Fourier transform of a signal, we will rarely use \@ref(eq:discrete-fourier). Instead, we will 
look up the Fourier transform of the signal in a table like Appendix \@ref(dtft) and use properties of the 
Fourier transform (as shown in Appendix \@ref(fourier-properties)).


## Essential Practice {-}

1. Calculate the Fourier transform $G(f)$ of the continuous-time signal 
\[ g(t) = \begin{cases} 1/3 & 0 < t < 3 \\ 0 & \text{otherwise} \end{cases}.  \]
Graph its magnitude $|G(f)|$ as a function of frequency.

2. Calculate the Fourier transform $G(f)$ of the discrete-time signal 
\[ g[n] = \delta[n] + 0.4 \delta[n-1] + 0.4 \delta[n+1]. \]
Graph $G(f)$. (The Fourier transform should be a real-valued function, so you do not need to take its magnitude.)

3. Which of the following terms fill in the blank? The Fourier transform $G(f)$ of a real-valued time signal $g(t)$ 
is necessarily ....

    a. real-valued: $G(f) \in \mathbb{R}$
    b. positive-valued: $G(f) > 0$
    c. symmetric: $G(-f) = G(f)$
    d. conjugate symmetric: $G(-f) = G^*(f)$
    
    (_Hint:_ Write out $G(f)$ and $G(-f)$ according to the definition \@ref(eq:fourier), and 
    use Euler's identity (Theorem \@ref(thm:euler)). Simplify as much as possible, using the facts that 
    $\cos(-\theta) = \cos(\theta)$ and $\sin(-\theta) = -\sin(\theta)$.)

4. Which of the following terms fill in the blank? The Fourier transform $G(f)$ of a real-valued and _symmetric_ 
time signal $g(t)$ is necessarily ....

    a. real-valued: $G(f) \in \mathbb{R}$
    b. positive-valued: $G(f) > 0$
    c. symmetric: $G(-f) = G(f)$
    d. conjugate symmetric: $G(-f) = G^*(f)$
    
    
