# LTI Filters in the Frequency Domain {#lti-frequency}

## Motivation {-}

In Lesson \@ref(lti-time), we studied what LTI filters do to stationary signals in the time 
domain. However, it is usually easier to analyze and understand LTI filters in the frequency 
domain, as we explore in this lesson.

## Theory {-}

```{theorem filter-theorem, name="Filter Theorem"}
Let $\{ X(t) \}$ be a stationary random process with power spectral density $S_X(f)$. 
If  $\{ X(t) \}$ is passed into a 
linear time-invariant filter with impulse response $h(t)$, then the output 
process $\{ Y(t) \}$ has power spectral density 
\begin{equation}
S_Y(f) = |H(f)|^2 \cdot S_X(f),
(\#eq:filter-theorem)
\end{equation}
where $H(f)$ is the frequency response of the filter (i.e., the Fourier transform of the 
                                        impulse response).
```

```{proof}
Theorem \@ref(thm:filter-stationary) provides the mean and autocovariance functions of $\{ Y(t) \}$. We can use these 
to obtain the autocorrelation function of $\{ Y(t) \}$:
\begin{align*}
R_Y(\tau) &= C_Y(\tau) + \mu_Y^2 \\
&= (h(-t) * h * C_X)(\tau) + \left(\mu_X \int_{-\infty}^\infty h(t)\,dt\right)^2 \\
&= (h(-t) * h * C_X)(\tau) + (h(-t) * h * \mu_X^2) \\
&= (h(-t) * h * R_X)(\tau).
\end{align*}
The second-to-last equality follows from the fact that the convolution of any function with a constant function 
($\mu_X^2$) is itself a constant function. So $h * \mu_X^2$ is the constant $\int_{-\infty}^\infty h(t)\,dt \mu_X^2$, 
and convolving this constant function with $h(-t)$ gives another factor of $\int_{-\infty}^\infty h(t)\,dt$.

Now, we know that $S_Y$, the power spectral density of $\{ Y(t) \}$, is just the Fourier transform of the 
autocorrelation function $R_Y$. Combining the representation of $R_Y$ above, with the Fourier properties from 
Appendix \@ref(fourier-properties), we obtain:
\begin{align*}
S_Y(f) &= \mathscr{F}[R_Y](f) \\
&= \mathscr{F}[h(-t) * h * R_X](f) \\
&= \mathscr{F}[h(-t)](f) \cdot \mathscr{F}[h](f) \cdot \mathscr{F}[R_X](f) & \text{(convolution property)} \\
&= H(-f) \cdot H(f) \cdot S_Y(f) & \text{(definitions of $H$ and $S_Y$, reversal property)} \\
&= H^*(f) \cdot H(f) \cdot S_X(f) & \text{($h$ is real, so $H$ is conjugate symmetric)} \\
&= |H(f)|^2 \cdot S_X(f),
\end{align*}
as we wanted to show.
```

```{example white-noise-lti-frequency, name="White Noise"}
Consider the white noise process $\{ Z[n] \}$ defined in Example \@ref(exm:white-noise), 
which consists of i.i.d. random variables with mean $\mu = E[Z[n]]$ and 
variance $\sigma^2 \overset{\text{def}}{=} \text{Var}[Z[n]]$. 

We showed in Example \@ref(exm:white-noise-psd) that its PSD is
\[ S_X(f) = \sigma^2 + \mu^2 \delta(f). \]

Let's pass $\{ X(t) \}$ through a filter with impulse response 
\[ h[n] = (-0.4)^n u[n]. \]
What is the PSD of the output $\{ Y(t) \}$?

By the Filter Theorem (Theorem \@ref(thm:filter-theorem)), we know that 
\[ S_Y(f) = |H(f)|^2 \cdot S_X(f). \]
So all we need to do is find $|H(f)|^2$.

By Table \@ref(dtft), the frequency response of the filter is 
\[ H(f) = \frac{1}{1 + 0.4e^{-j2\pi f}} \]
so 
\begin{align*}
|H(f)|^2 &= H(f) \cdot H^*(f) \\
&= \frac{1}{1 + 0.4e^{-j2\pi f}} \cdot \frac{1}{1 + 0.4e^{j2\pi f}} \\
&= \frac{1}{1 + 0.4 (e^{j2\pi f} + e^{-j2\pi f}) + (0.4)^2} \\
&= \frac{1}{1.16 + 0.8\cos(2\pi f)}.
\end{align*}

Now, $S_Y(f)$ is just the product of this and $S_X(f)$.
\begin{align*} 
S_Y(f) &= \frac{1}{1.16 + 0.8\cos(2\pi f)} \cdot (\sigma^2 + \mu^2 \delta(f)) \\
&= \frac{\sigma^2}{1.16 + 0.8\cos(2\pi f)} + \frac{\mu^2}{1.16 + 0.8\cos(2\pi \cdot 0)} \delta(f).
\end{align*}
The last equality follows from the fact that $\delta(f) = 0$ everywhere except $f=0$, 
so we can substitute $f=0$ into the coefficient.

The graphs below show:

- the PSD of the input $S_X(f)$ (in red), 
- the squared magnitude of the frequency response $|H(f)|^2$ (in blue), and 
- the PSD of the output $S_Y(f)$ (in orange), 

for $\mu=-1.5$ and $\sigma^2 = 1.7$. Notice that the filter $h$ is a highpass filter, passing 
high frequencies (i.e., near the Nyquist limit of 0.5) and attenuating low frequencies.
```

```{r, echo=FALSE, fig.show = "hold", fig.align = "default", fig.asp=1.3}
par(mfrow=c(2, 1), mgp=c(1, 1, 0))

fs <- seq(-0.5, 0.5, .001)

plot(0, 1.5^2,
     xaxt="n", yaxt="n", bty="n",
     col="red", pch=19,
     xlim=c(-0.5, 0.5), ylim=c(0, 4.5),
     xlab="Frequency (cycles/sample)", 
     ylab="Power Spectral Density")
axis(1, pos=0, at=(-5:5) / 10)
axis(2, pos=0, at=(1:8) / 2)
lines(c(-0.5, 0.5), c(1.7, 1.7), type="l", col="red")
lines(c(0, 0), c(0, 1.5^2), lwd=2, col="red")
lines(fs, 1 / (1.16 + 0.8 * cos(2 * pi * fs)), col="blue")

plot(0, 1.148,
     xaxt="n", yaxt="n", bty="n",
     col="orange", pch=19,
     xlim=c(-0.5, 0.5), ylim=c(0, 4.5),
     xlab="Frequency (cycles/sample)", 
     ylab="Power Spectral Density")
axis(1, pos=0, at=(-5:5) / 10)
axis(2, pos=0, at=(1:8) / 2)
lines(c(0, 0), c(0, 1.148), lwd=2, col="orange")
lines(fs, 1.7 / (1.16 + 0.8 * cos(2 * pi * fs)), col="orange")
```

We can calculate the expected power in the output signal $\{ Y[n] \}$ by integrating 
the PSD over "all" frequencies. (This is the easiest way, since 
the autocorrelation function $R_Y(\tau)$ is not available.) Because this is a discrete-time 
signal, we only integrate over frequencies below the Nyquist limit, $|f| < 0.5$.
\begin{align*}
\int_{-0.5}^{0.5} S_Y(f)\,df &= \int_{-0.5}^{0.5} \frac{\sigma^2}{1.16 + 0.8\cos(2\pi f)} \,df + \int_{-0.5}^{0.5} \frac{\mu^2}{1.96} \delta(f)\,df \\
&\approx 1.1905 \sigma^2 + \frac{\mu^2}{1.96}.
\end{align*}

```{example gaussian-process-lti}
Consider the process $\{X(t)\}$ from Example \@ref(exm:gaussian-process) with 
autocorrelation function 
\[ R_X(\tau) = 5 e^{-\tau^2 / 3} + 4. \]
We showed in Example \@ref(exm:gaussian-process-psd) that its power spectral density is 
\begin{align*}
S_X(f) &=  5 \sqrt{\frac{\pi}{3}} e^{-\pi^2 f^2 / 3}  + 4 \delta(f)
\end{align*}
Now, suppose we pass $\{ X(t) \}$ through an LTI filter with impulse response 
\[ h(t) = 1.8\text{sinc}(1.2f). \]
Using the table in Appendix \@ref(ctft), along with the scaling property, we see that the frequency response is 
\[ H(f) = 1.8\frac{1}{1.2} \text{rect}(\frac{f}{1.2}) = \begin{cases} 1.5 & |f| < 0.6 \\ 0 & \text{otherwise} \end{cases}. \]
The filter $h$ is a ideal lowpass filter because it perfectly preserves frequencies below $0.6$ Hz and 
perfectly removes frequencies above $0.6$ Hz. It is "ideal" because the impulse response $h(t)$ is 
not implementable.

Now, we have to calculate the _squared_ magnitude of the impulse response for the filter theorem:
\[ |H(f)|^2 = \begin{cases} 1.5^2 & |f| < 0.6 \\ 0 & \text{otherwise} \end{cases}. \]

$S_X(f)$ and $|H(f)|^2$ are graphed below.
```

```{r, echo=FALSE, fig.show = "hold", fig.align = "default"}
plot(0, 4,
     xaxt="n", yaxt="n", bty="n",
     col="red", pch=19,
     xlim=c(-2, 2), ylim=c(0, 11.5),
     xlab="Frequency (Hz)", 
     ylab="Power Spectral Density")
axis(1, pos=0, at=(-2:2))
axis(2, pos=0, at=seq(2, 10, 2))

fs <- seq(-2, 2, 0.01)
lines(fs, 5 * sqrt(pi / 3) * exp(-pi^2 * fs^2 / 3), type="l", col="red")
lines(c(0, 0), c(0, 4), lwd=2, col="red")
lines(c(-2, -.6, -.6, .6, .6, 2), c(0, 0, 1.5^2, 1.5^2, 0, 0), lwd=2, col="blue")
```

By the Filter Theorem (Theorem \@ref(thm:filter-theorem)), the PSD of the output is 
\[ S_Y(f) = |H(f)|^2 \cdot S_X(f) = \begin{cases} 1.5^2 \cdot 5 \sqrt{\frac{\pi}{3}} e^{-\pi^2 f^2 / 3}  + 1.5^2 \cdot 4 \delta(f) & |f| < 0.6 \\ 0 & \text{otherwise} \end{cases} \]
Although the formula for $S_Y(f)$ is messy, it is quite straightforward to understand what is going on if we just 
graph it.

```{r, echo=FALSE, fig.show = "hold", fig.align = "default"}
plot(0, 1.5^2 * 4,
     xaxt="n", yaxt="n", bty="n",
     col="orange", pch=19,
     xlim=c(-2, 2), ylim=c(0, 11.5),
     xlab="Frequency (Hz)", 
     ylab="Power Spectral Density")
axis(1, pos=0, at=(-2:2))
axis(2, pos=0, at=seq(2, 10, 2))

fs <- seq(-0.6, 0.6, 0.01)
lines(fs, 1.5^2 * 5 * sqrt(pi / 3) * exp(-pi^2 * fs^2 / 3), col="orange")
lines(c(-2, -0.6, -0.6), c(0, 0, 1.5^2 * 5 * sqrt(pi / 3) * exp(-pi^2 * 0.6^2 / 3)), col="orange")
lines(c(2, 0.6, 0.6), c(0, 0, 1.5^2 * 5 * sqrt(pi / 3) * exp(-pi^2 * 0.6^2 / 3)), col="orange")
lines(c(0, 0), c(0, 1.5^2 * 4), lwd=2, col="orange")
```

Frequencies below 0.6 Hz are passed (with a gain of $1.5^2 = 2.25$), while 
frequencies above 0.6 Hz are rejected completely.

The Filter Theorem allows us to prove a fact about power spectral densities that we have 
assumed until now.

```{theorem psd-band-integral, name="Expected Power in a Frequency Band"}
The integral of the PSD $S_X(f)$ over any frequency band $a < |f| < b$ equals the expected power 
in that frequency band.
```

```{r, echo=FALSE, engine='tikz', out.width='70%', fig.ext='png', fig.align='center', engine.opts = list(template = "tikz_template.tex")}
	\begin{tikzpicture}
	
	\node (x) at (0, -0.1) {$X(t)$};
	\node[draw,blue] (h) at (2, -0.1) {$H(f)$};
	\node (y) at (4, -0.1) {$Y(t)$};

  \draw[->,gray] (1.0, 0.5) -- (3.0, 0.5);
  \draw[->,gray] (2, 0.4) -- (2, 1.2);
  \node[anchor=west] at (2.95, 0.5) {{\tiny $f$}};

  \draw[blue,thick] (1.0, 0.5) -- (1.4, 0.5) -- (1.4, 0.8) -- (1.8, 0.8) -- (1.8, 0.5) -- (2.2, 0.5) -- (2.2, 0.8) -- (2.6, 0.8) -- (2.6, 0.5) -- (2.95, 0.5);

  \node[anchor=north] at (1.4, 0.55) {{\tiny $-b$}};
  \node[anchor=north] at (1.8, 0.55) {{\tiny $-a$}};
  \node[anchor=north] at (2.2, 0.5) {{\tiny $a$}};
  \node[anchor=north] at (2.6, 0.55) {{\tiny $b$}};
	
	\draw[->] (x) -- (h);
	\draw[->] (h) -- (y);	
	
	\end{tikzpicture}
```

```{proof}
To calculate the expected power in a signal $\{ X(t) \}$ over the frequency band $a < |f| < b$, 
we imagine passing $\{ X(t) \}$ through the ideal bandpass filter 
\[ H(f) = \begin{cases} 1 & a < |f| < b \\ 0 & \text{otherwise} \end{cases}. \]
The output signal $\{ Y(t) \}$ is then just a version of $\{ X(t) \}$, with all frequency content 
outside $a < |f| < b$ removed.

To obtain the expected power in this frequency band, we integrate the PSD $S_Y(f)$ of the output over all frequencies. 
For continuous-time signals, this looks like  
\[ \text{Expected power between $a$ and $b$} = \int_{-\infty}^\infty S_Y(f)\,df. \]
(For discrete-time signals, the limits of the integral are $-0.5$ and $0.5$.)

But by the Filter Theorem (Theorem \@ref(thm:filter-theorem)), we have 
\[ \int_{-\infty}^\infty S_Y(f)\,df = \int_{-\infty}^\infty |H(f)|^2 S_X(f)\,df = \int_{a < |f| < b} S_X(f)\,df, \]
since $|H(f)|^2$ is 0 outside $a < |f| < b$. This shows that we can obtain the expected power over any frequency 
band by integrating the PSD over the appropriate frequencies.
```

## Essential Practice {-}

For these questions, you may want to refer to the power spectral densities that 
you calculated in Lesson \@ref(psd).

1. Radioactive particles hit a Geiger counter according to a Poisson process 
at a rate of $\lambda=0.8$ particles per second. Let $\{ N(t); t \geq 0 \}$ represent this Poisson process.

    Define the new process $\{ D(t); t \geq 3 \}$ by 
    \[ D(t) = N(t) - N(t - 3). \]
    This process represents the number of particles that hit the Geiger counter in the 
    last 3 seconds. 
    
    Suppose $\{ D(t) \}$ is passed through an LTI filter with impulse response 
    \[ h(t) = 3\text{sinc}(2t). \]
    
    What is the power spectral density of the output? What is the expected power in the output?
    
2. Consider the moving average process $\{ X[n] \}$ of Example \@ref(exm:ma1), defined by 
\[ X[n] = 0.5 Z[n] + 0.5 Z[n-1], \]
where $\{ Z[n] \}$ is a sequence of i.i.d. standard normal random variables. 

    In Lesson \@ref(lti-time), you expressed $\{ X[n] \}$ as white noise passed through an LTI filter 
    and calculated the impulse response of this filter. Use this impulse response and the 
    PSD of white noise to calculate the PSD of this moving average process. Check 
    that your answer agrees with the one you obtained in Lesson \@ref(psd).

3. Let $\{ X(t) \}$ be a continuous-time random process with mean function 
$\mu_X(t) = -2$ and autocovariance function $C_X(s, t) = 5 e^{-0.4|s - t|}$. 

    If $\{X(t)\}$ is passed through an LTI filter with impulse response 
    \[ h(t) = \begin{cases} 4e^{-2t} & t \geq 0 \\ 0 & \text{otherwise} \end{cases}, \]
    what is the expected power in the output?
