# Normal Distribution {#normal}

## Motivation {-}

<iframe width="560" height="315" src="https://www.youtube.com/embed/sScArppyv_g" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

The normal distribution is the most important distribution in all of probability and statistics. 
It is ubiquitous in the real world, as the above video demonstrates. 

The normal distribution goes by a few other names:

- Gaussian distribution (especially common in engineering)
- the bell curve (in everyday life)

We will discuss the normal distribution in two stages:

1. First, we will discuss the standard normal distribution.
2. Then, we will discuss the general normal distribution, which has the same shape as the 
standard normal distribution, but with a different center and scale.

## Standard Normal Distribution {-}

<iframe width="560" height="315" src="https://www.youtube.com/embed/hhuvWCDgkEk" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

```{definition standard-normal, name="Standard Normal Distribution"}
A random variable $Z$ is said to follow a standard normal distribution if 
its p.d.f. is 
\begin{equation}
f(z) = \frac{1}{\sqrt{2\pi}} e^{-z^2 / 2}.
(\#eq:standard-normal-pdf)
\end{equation}
The p.d.f. is non-zero for all real numbers $-\infty < z < \infty$. It is conventional to use the 
letter $Z$ for a standard normal random variable.

The c.d.f. of the standard normal distribution is 
\begin{equation}
\Phi(x) \overset{\text{def}}{=} \int_{-\infty}^x f(t)\,dt. 
(\#eq:standard-normal-cdf)
\end{equation}
Note that it is conventional to use the Greek letter $\Phi$ ("phi") for the standard normal c.d.f., instead of $F$. 
There is no closed-form formula for $\Phi$, so we often leave our answers in terms of $\Phi$. If a numerical 
value is required, we have a few options:
  
  - We can use numerical integration.
  - We can use software that calculates normal probabilities.
  - We can use tables of the normal c.d.f., like [this one](https://web.calpoly.edu/~dsun09/tables/normal_cdf_table.pdf).
```

The figure below shows the p.d.f. and c.d.f. of a standard normal distribution.

```{r standard-normal-graph, echo=FALSE, fig.show = "hold", fig.align = "center",fig.cap="PDF and CDF of the Standard Normal Distribution", fig.asp=1.2}
par(mfrow=c(2, 1), mgp=c(1, 1, 0))

t <- seq(-3.5, 3.5, by=0.01)
plot(t, dnorm(t), type="l", lwd=3, col="blue", 
     xaxt="n", yaxt="n", bty="n",
     xlab="z", ylab="Probability Density Function f(z)", 
     xlim=c(-3.5, 3.5), ylim=c(-0.05, 0.45))
polygon(c(-3.5, t[t <= 0.5], 0.5), c(0, dnorm(t[t <= 0.5]), 0), col=rgb(1, 0, 0, .5), border=NA)
axis(1, pos=0)
axis(2, pos=0)

plot(t, pnorm(t), type="l", lwd=3, col="blue",
     xaxt="n", yaxt="n", bty="n",
     xlab="x", ylab="Cumulative Distribution Function F(x)", 
     xlim=c(-3.5, 3.5), ylim=c(-0.05, 1.05))
axis(1, pos=0)
axis(2, pos=0)
lines(c(0.5, 0.5, 0), c(0, pnorm(0.5), pnorm(0.5)), lty=2, col=rgb(1, 0, 0))
points(0.5, pnorm(0.5), col=rgb(1, 0, 0, .5), pch=19)
```

The Colab below shows how to use software to calculate probabilities involving a standard normal distribution.

<script src="https://gist.github.com/dlsun/d3d113724f70143fbe004dfc555533b5.js"></script>

```{theorem, name="Expected Value and Variance of the Standard Normal"}
Let $Z$ be a standard normal random variable. Then:
  \begin{align*}
E[Z] &= 0 & \text{Var}[Z] &= 1
\end{align*}

```{proof}
\begin{align*}
E[Z] &= \int_{-\infty}^\infty z \cdot \frac{1}{\sqrt{2\pi}} e^{-z^2/2} \\
&= -\frac{1}{\sqrt{2\pi}} e^{-z^2/2} \Big|_{-\infty}^\infty \\
&= (-0 + 0) \\
&= 0 \\
\text{Var}[Z] &= E[Z^2] - E[Z]^2 \\
&= \int_{-\infty}^\infty z^2 \cdot \frac{1}{\sqrt{2\pi}} e^{-z^2/2} - 0^2 \\
&= 1
\end{align*}
```


## (General) Normal Distribution {-}

```{definition normal, name="Normal Distribution"}
A random variable $X$ is said to follow a $\text{Normal}(\mu, \sigma)$ 
  distribution if it can be expressed as 
\begin{equation}
X = \mu + \sigma Z, 
(\#eq:loc-scale-normal)
\end{equation}
where $Z$ is a standard normal random variable. 

Its p.d.f. is 
\begin{equation}
f(x) = \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{1}{2} ( x - \mu)^2}. 
(\#eq:normal-pdf)
\end{equation}
To derive this, apply Theorems \@ref(thm:scale-transform) and 
\@ref(thm:shift-transform) to \@ref(eq:standard-normal-pdf).
However, we will rarely use this p.d.f. Instead, we typically use the 
representation \@ref(eq:loc-scale-normal) to convert the distribution 
to a standard normal before doing any calculations.

Note that the standard normal distribution is the $\text{Normal}(\mu=0, \sigma=1)$ distribution.
```

Shown below are the p.d.f.s and c.d.f.s of normal distributions for different values of 
$\mu$ and $\sigma$. Notice that $\mu$ controls where the bell is centered, while $\sigma$ 
controls how wide it is.

```{r normal-graph, echo=FALSE, fig.show = "hold", fig.align = "center",fig.cap="PDF and CDF of the Normal Distribution", fig.asp=1.2}
par(mfrow=c(2, 1), mgp=c(1, 1, 0))

t <- seq(-4.5, 4.5, by=0.01)
plot(t, dnorm(t), type="l", lwd=3, col="blue", 
     xaxt="n", yaxt="n", bty="n",
     xlab="x", ylab="Probability Density Function f(x)", 
     xlim=c(-4.5, 4.5), ylim=c(-0.05, 0.8))
axis(1, pos=0)
axis(2, pos=0)
lines(t, dnorm(t, 2, 0.5), type="l", lwd=3, col="red")
lines(t, dnorm(t, 0, 2), type="l", lwd=3, col="orange")
legend(-4, 0.8, 
       c(expression(paste(mu, " = 0, ", sigma, " = 1")),
         expression(paste(mu, " = 2, ", sigma, " = 0.5")), 
         expression(paste(mu, " = 0, ", sigma, " = 2"))),
       lwd=3, col=c("blue", "red", "orange"))

plot(t, pnorm(t), type="l", lwd=3, col="blue",
     xaxt="n", yaxt="n", bty="n",
     xlab="x", ylab="Cumulative Distribution Function F(x)", 
     xlim=c(-4.5, 4.5), ylim=c(-0.05, 1.05))
lines(t, pnorm(t, 2, 0.5), type="l", lwd=3, col="red")
lines(t, pnorm(t, 0, 2), type="l", lwd=3, col="orange")
axis(1, pos=0)
axis(2, pos=0)
legend(-4, 1, 
       c(expression(paste(mu, " = 0, ", sigma, " = 1")),
         expression(paste(mu, " = 2, ", sigma, " = 0.5")), 
         expression(paste(mu, " = 0, ", sigma, " = 2"))),
       lwd=3, col=c("blue", "red", "orange"))
```

Note that \@ref(eq:loc-scale-normal) implies that for a normal random variable,
\begin{equation}
Z = \frac{X - \mu}{\sigma}
(\#eq:standardize)
\end{equation}
follows a standard normal distribution. This process is known as **standardization**. 
So to calculate probabilities for a normal random variable, we can standardize the 
random variable so that it follows a _standard_ normal distribution and calculate 
probabilities from the standard normal distribution.

The next theorem formally establishes that $\mu$ controls the center and 
$\sigma$ the spread of the normal distribution.

```{theorem, name="Expected Value and Variance of the Normal"}
Let $X$ be a $\text{Normal}(\mu, \sigma)$ random variable. Then:
  \begin{align*}
E[X] &= \mu & \text{Var}[X] &= \sigma^2
\end{align*}
so that $\mu$ and $\sigma$ represent the mean and standard deviation, respectively.
```{proof}
The result follows by the representation \@ref(eq:loc-scale-normal) and 
properties of expected value and variance:
\begin{align*}
E[X] &= E[\mu + \sigma Z] \\
&= \mu + \sigma E[Z] \\
&= \mu \\
\text{Var}[X] &= \text{Var}[\mu + \sigma Z] \\
&= \sigma^2 \text{Var}[Z] \\
&= \sigma^2
\end{align*}
```

The case study below shows how to calculate probabilities from a normal distribution, 
using a combination of algebra and software.

<script src="https://gist.github.com/dlsun/6da723d1b80742172f8440c89df87246.js"></script>


## Essential Practice {-}

1. Based on extensive data from an urban freeway near Toronto, Canada, "it is assumed that free
speeds can best be represented by a normal distribution" ("Impact of Driver Compliance on the
Safety and Operational Impacts of Freeway Variable Speed Limit Systems," _J. of Transp. Engr._,
2011: 260–268). The mean and standard deviation reported in the article were 119 km/h and
13.1 km/h, respectively.

    a. What is the probability that the speed of a randomly selected vehicle is between 100 and
120 km/h?
    b. What speed characterizes the fastest 10% of all speeds?
    c. If five vehicles are randomly and independently selected, what is the probability that at
least one car is traveling under the posted speed limit of 100 km/h?

2. Daily highs in San Luis Obispo in August are approximately normally distributed with a mean of 
$76.9^\circ\textrm{F}$. The temperature exceeds 100 degrees Fahrenheit on about 1.5% 
of August days.

    a. What can you say about the standard deviation?
    b. Suppose the mean increases by 2 degrees Fahrenheit. By what _factor_ will the 
    percentage of 100-degree days increase?
    
3. Suppose that the wrapper of a certain candy bar lists its weight as 2.13 ounces. Naturally, the
weights of individual bars vary somewhat. Suppose that the actual weights of these candy bars
vary according to a normal distribution with mean $\mu = 2.20$ ounces and standard deviation 
$\sigma = 0.04$ ounces.

    a. What proportion of candy bars weigh less than the advertised weight?
    b. If the weights of candy bar are independent, what is the expected number of candy bars 
    before you encounter one that weights less than the advertised weight?
    c. If the manufacturer decides that it’s unacceptable to have so many candy bars weigh less than
the advertised weight, they might want to adjust the production process so that only 1 candy bar
in 1000 weighs less than advertised. What should the mean of the actual weights be (assuming
that the standard deviation of the weights remains 0.04 ounces)? Is this more or less than before?
Why does this makes sense?
    d. If the manufacturer does not want to add weight to the candy bars (because this costs money),
an alternative is to adjust the SD of the weights in the production process. If the mean weight
remains at 2.20 ounces but only 1 candy bar in 1000 weighs less than the advertised weight, how
small does the standard deviation of the weights need to be? Is this smaller or larger than
before? Why does this makes sense?

4. Let $Z$ be a standard normal random variable. Derive the p.d.f. of $X = e^Z$. Sketch this p.d.f. 

    (This distribution has a name: the log-normal distribution. This is a popular distribution for modeling 
    random variables with long right tails, such as income. Hopefully, you can appreciate this if you sketch 
    the p.d.f.)
