# Power Spectral Density {#psd}

## Motivation {-}

In Lesson \@ref(power), we saw that the expected power of a stationary 
random signal $\{ X(t) \}$ could be calculated by evaluating the autocorrelation 
function $R_X(\tau)$ at a difference of $\tau = 0$. But how is this expected power 
distributed across different frequencies? The answer also involves the 
autocorrelation function.

## Theory {-}

```{definition psd, name="Power Spectral Density"}
The **power spectral density** (or PSD, for short) $S_X(f)$ of a stationary 
random process $\{ X(t) \}$ is the Fourier transform of the autocorrelation function 
$R_X(\tau)$. (Note: Because the process is stationary, the autocorrelation  
              only depends on the difference $\tau = s - t$.)

That is, the autocorrelation function and the power spectral density are Fourier pairs.
\begin{equation} 
R_X(\tau) \overset{\mathscr{F}}{\longleftrightarrow} S_X(f). 
(\#eq:autocorrelation-psd)
\end{equation}
```

In Lesson \@ref(lti-frequency), we will motivate why it makes 
sense to interpret the Fourier transform of the autocorrelation 
function as the distribution of power over frequency. For now, 
we take \@ref(eq:autocorrelation-psd) to be the 
_definition_ of the power spectral density.

```{example white-noise-psd, name="White Noise"}
Consider the white noise process $\{ Z[n] \}$ defined in Example \@ref(exm:white-noise), 
which consists of i.i.d. random variables with mean $\mu = E[Z[n]]$ and 
variance $\sigma^2 \overset{\text{def}}{=} \text{Var}[Z[n]]$. 

This process is stationary, with autocorrelation function 
\[ R_X[k] = \sigma^2\delta[k] + \mu^2. \]
The power spectral density is its Fourier transform. Using 
linearity and 
the Discrete-Time Fourier Transform table in Appendix \@ref(dtft), we see that 
\[ S_X(f) = \sigma^2 + \mu^2 \delta(f). \]

This PSD is graphed below for $\mu=-1.5$ and 
$\sigma^2 = 1.7$. Note that this power spectral density is only defined 
for frequencies below the Nyquist limit: $|f| < 0.5$. We see that 
white noise has equal power at all frequencies, except at 0 Hz (DC).
```

```{r, echo=FALSE, fig.show = "hold", fig.align = "default"}
plot(0, 1.5^2,
     xaxt="n", yaxt="n", bty="n",
     col="red", pch=19,
     xlim=c(-0.5, 0.5), ylim=c(0, 2.5),
     xlab="Frequency (cycles/sample)", 
     ylab="Power Spectral Density")
axis(1, pos=0, at=(-5:5) / 10)
axis(2, pos=0, at=(1:5) / 2)
lines(c(-0.5, 0.5), c(1.7, 1.7), type="l", col="red")
lines(c(0, 0), c(0, 1.5^2), lwd=2, col="red")
```

One way to check that we calculated the PSD correctly is to integrate 
it over "all" frequencies and making sure it matches the expected power 
we calculated in Lesson \@ref(power). Because this is a discrete-time signal, 
we only integrate over frequencies below the Nyquist limit, $|f| < 0.5$.
\[ \int_{-0.5}^{0.5} S_X(f)\,df = \int_{-0.5}^{0.5} \sigma^2\,df + \mu^2 \int_{-0.5}^{0.5} \delta(f)\,df = \sigma^2 + \mu^2  \]

The power in this signal below 0.2 cycles/sample can be calculated 
by integrating the PSD, 
remembering to include both positive _and_ negative frequencies.

\[ \int_{-0.2}^{0.2} S_X(f)\,df = \int_{-0.2}^{0.2} \sigma^2\,df + \mu^2 \int_{-0.2}^{0.2} \delta(f)\,df = 0.4\sigma^2 + \mu^2. \]

```{example gaussian-process-psd}
Consider the process $\{X(t)\}$ from Example \@ref(exm:gaussian-process). 
This process is stationary, with autocorrelation function
\[ R_X(\tau) = 5 e^{-\tau^2 / 3} + 4. \]
The power spectral density is its Fourier transform. Using linearity, scaling, and 
the Continuous-Time Fourier Transform table in Appendix \@ref(ctft), we see that 
\begin{align*}
S_X(f) &= \mathscr{F}[5 e^{-3 \tau^2} + 4] \\
&= 5 \mathscr{F}[\underbrace{e^{-3 \tau^2}}_{e^{-\pi \left(\sqrt{\frac{3}{\pi}}\tau\right)^2}}] + 4 \mathscr{F}[1] & \text{(linearity)} \\
&= 5 \sqrt{\frac{\pi}{3}} e^{-\pi (\sqrt{\frac{\pi}{3}} f)^2}  + 4 \delta(f) & \text{(scaling by $a = \sqrt{\frac{3}{\pi}}$)}
\end{align*}
Notice how we massaged $e^{-3\tau^2}$ into the form 
$e^{-\pi (a\tau)^2}$, which is just a scale transform of one of the 
transforms in Appendix \@ref(ctft).

This power spectral density is graphed below. We see that this process has 
more power at the lower frequencies than at the higher frequencies.
```

```{r, echo=FALSE, fig.show = "hold", fig.align = "default"}
plot(0, 4,
     xaxt="n", yaxt="n", bty="n",
     col="red", pch=19,
     xlim=c(-2, 2), ylim=c(0, 5),
     xlab="Frequency (Hz)", 
     ylab="Power Spectral Density")
axis(1, pos=0, at=(-2:2))
axis(2, pos=0, at=(1:5))

fs <- seq(-2, 2, 0.01)
lines(fs, 5 * sqrt(pi / 3) * exp(-pi^2 * fs^2 / 3), type="l", col="red")
lines(c(0, 0), c(0, 4), lwd=2, col="red")
```

One way to check that we calculated the PSD correctly is to integrate 
it over all frequencies and making sure it matches the expected power 
we calculated in Lesson \@ref(power).
\[ \int_{-\infty}^{\infty} S_X(f)\,df = \int_{-\infty}^{\infty} 5 \sqrt{\frac{\pi}{3}} e^{-\pi^2 f^2 / 3}\,df + 4 \int_{-\infty}^{\infty} \delta(f)\,df = 5 + 4 = 9. \]

The power in this signal below above 1 Hz can be calculated 
by integrating the PSD, 
remembering to include both positive _and_ negative frequencies. Since the 
PSD is symmetric, the easiest way to do this is to double the integral over 
$1 < f < \infty$.

\[ 2 \int_{1}^{\infty} S_X(f)\,df = 2  \int_1^{\infty} 5 \sqrt{\frac{\pi}{3}} e^{-\pi^2 f^2 / 3}\,df = .0515. \]


## Essential Practice {-}

For these questions, you may want to refer to the autocorrelation function 
you calculated in Lesson \@ref(autocorrelation).

1. Radioactive particles hit a Geiger counter according to a Poisson process 
at a rate of $\lambda=0.8$ particles per second. Let $\{ N(t); t \geq 0 \}$ represent this Poisson process.

    Define the new process $\{ D(t); t \geq 3 \}$ by 
    \[ D(t) = N(t) - N(t - 3). \]
    This process represents the number of particles that hit the Geiger counter in the 
    last 3 seconds. Calculate the PSD of $\{ D(t); t \geq 3 \}$. What is the 
    expected power above 1 Hz?
 
2. Consider the moving average process $\{ X[n] \}$ of Example \@ref(exm:ma1), defined by 
\[ X[n] = 0.5 Z[n] + 0.5 Z[n-1], \]
where $\{ Z[n] \}$ is a sequence of i.i.d. standard normal random variables. 
Calculate the PSD of $\{ X[n] \}$. What is the expected power below 0.1 cycles per 
sample?
    
3. Let $\Theta$ be a $\text{Uniform}(a=-\pi, b=\pi)$ random variable, and let $f$ be a constant. Define the random phase process $\{ X(t) \}$ by
\[ X(t) = \cos(2\pi f t + \Theta). \]
Calculate the PSD of $\{ X(t) \}$.

4. Let $\{ X(t) \}$ be a random process with mean function 
$\mu_X(t) = -2$ and autocovariance function $C_X(s, t) = 5 e^{-0.4|s - t|}$. 
Calculate the PSD of $\{ X(t) \}$. What is the expected power below 3 Hz?
